\Chapter{A New Chapter}{Philippa Harper, née  Court}

After nearly 30 years outside the family, when it was occupied by Betty Small and her husband, my husband Bryan and I bought the house back in March 2014. It was a key turning point in our lives. The whole family had missed the place greatly during the whole time.

I always felt hugely lucky to grow up in such a wonderful spot as Washford. We all loved the beauty of the home, the shop, the Post Office, the outhouses, the garden, the orchard, and were lucky to grow up when there was such a strong sense of community around us, with lots of kindness, good humour, and wonderful neighbours particularly in the family of the Bakers across the road.

Since then we’ve spent four years of hard work doing up the house and garden, as far as we can in keeping with its old spirit, and with respect to the endless patience, hard work and vision of our inspirational grandparents who built the place, and my hard-working parents who kept it going. We want to continue to care for Washford as best as we can. As everything grows so well, it is lucky that we both enjoy grass cutting and weeding!

We hope that the family will continue enjoying and caring for this space for a long time to come. Our own boys, Jonny, Chris and Mikey, have got to know and care for it too. We are really looking forward to this summer, when our youngest son, Mikey and his fiancée Agi will have a party here to celebrate the blessing of their wedding, and we are thrilled as the story of the house moves forward to the next generation.

Long Live Washford!\\


Washford, West Somerset\\
Spring, 2018
